---
title: "Air Quality Exam Project"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# 0. Data Preparation

Loading of the Data
```{r}
data <- read.table("AirQualityUCI.csv", header = TRUE, sep = ";",dec = ",")

```

Construct dataframes with relevant data.
```{r}
dataframeco <- data.frame(
  id = c(0:9470),
  date = data$Date,
  dayssince = c(0:0),
  time = data$Time,
  cogt = data$CO.GT,
  temperature = data$T
)

dataframenox <- data.frame(
  id = c(0:9470),
  date = data$Date,
  dayssince = c(0:0),
  time = data$Time,
  noxgt = data$NOx.GT.,
  temperature = data$T
)

dataframeno2 <- data.frame(
  id = c(0:9470),
  date = data$Date,
  time = data$Time,
  no2gt = data$NO2.GT.,
  temperature = data$T
)
```

Remove all the data with Not Assigned values.

```{r}
 dataframeco <- dataframeco[dataframeco$time >= 0, ] 
 dataframenox <- dataframenox[dataframenox$time >= 0, ] 
 dataframeno2 <- dataframeno2[dataframeno2$time >= 0, ] 
 
 dataframeco <- dataframeco[dataframeco$cogt != -200.0, ]
 dataframenox <- dataframenox[dataframenox$noxgt != -200.0, ]
 dataframeno2 <- dataframeno2[dataframeno2$no2gt != -200.0, ]
 
 dataframeco <- dataframeco[dataframeco$temperature != -200.0, ]
 dataframenox <- dataframenox[dataframenox$temperature != -200.0, ]
 dataframeno2 <- dataframeno2[dataframeno2$temperature != -200.0, ]
```

Derive Additional Variable "Days Since". That will denote days since data observation begun

```{r}
 date1string <- "10/03/2004"
   date1 <-  strptime(as.character(date1string), "%d/%m/%Y")
   results <- numeric(nrow(dataframeco))
   for (i in 1:nrow(dataframeco)) {
     date2string <- dataframeco$date[i]
     date2 <-  strptime(as.character(date2string), "%d/%m/%Y")
     results[i] <-  difftime(date2,date1, units = "days")
   }
   dataframeco$dayssince <- round(results, digits = 0)
   
   results <- numeric(nrow(dataframenox))
   for (i in 1:nrow(dataframenox)) {
     date2string <- dataframenox$date[i]
     date2 <-  strptime(as.character(date2string), "%d/%m/%Y")
     results[i] <-  difftime(date2,date1, units = "days")
   }
   dataframenox$dayssince <- round(results, digits = 0)
```

Restructure time variable

```{r}
dataframeco$time <- gsub(".00.00", "", dataframeco$time)
dataframeco$time <- as.numeric(dataframeco$time)
dataframenox$time <- gsub(".00.00", "", dataframenox$time)
dataframenox$time <- as.numeric(dataframenox$time)
dataframeno2$time <- gsub(".00.00", "", dataframeno2$time)
dataframeno2$time <- as.numeric(dataframeno2$time)

```
  
Update Id  

```{r}
    ids <- numeric(nrow(dataframeco))
    x <- -1
    for (i in 1:nrow(dataframeco)) {
      x = x + 1
      ids[i] <-  x
    }
    dataframeco$id <- ids
    
    ids <- numeric(nrow(dataframenox))
    y <- -1
    for (i in 1:nrow(dataframenox)) {
      y = y + 1
      ids[i] <-  y
    }
    dataframenox$id <- ids
```

 
#  1. Quantitative Exploratory Analysis
  
  Selected Variables: NOx (Kvælstofoxider), CO (Carbon Monoxide) Temperature
   
```{r}

summary(dataframenox$nox)
var(dataframenox$nox)
sd(dataframenox$nox)

```

There range is very big. As the 3rd quantile is 321 and the max is 1479, there are relatively few observations in the more extremely high values. And as the median is lower than the mean that also means that less observations are in the lower half than in the upper one. The difference between minimum-1st quantile, 1st quantile-median, and median-3rd quantile are, however, all relatively similar. All in all, the distribution seems to be quite left-skewed (positive)

```{r}

summary(dataframeco$co)
var(dataframeco$co)
sd(dataframeco$co)

```

Interestingly a lot smaller range than for NOx. The situation with max is however very similar to that of Nox, as the difference between 3rd quantile and max is a lot bigger than all other quantile differences. The mean is also higher than the median. Again, this is a left-skewed (positive) distribution. 


```{r}
summary(dataframenox$temperature)
var(dataframenox$temperature)
sd(dataframenox$temperature)
```
This should be left-skewed again, but less obviously. The difference between 3rd quantile and maximum is still the biggest of all, but the range between minimum and 1st quantile is also quite big. Meaning, the distribution is still left-skewed, but a lot more symmetrical than for NOx and CO. Most observations are close to the median.

```{r}

summary(dataframeco$temperature)
var(dataframeco$temperature)
sd(dataframeco$temperature)

```


#  2. Visual Exploratory Analysis



```{r}

df_list <- list(
  co = list(df = dataframeco, pollutant = "cogt", temp = "temperature", name = "CO"),
  nox = list(df = dataframenox, pollutant = "noxgt", temp = "temperature", name = "NOx")
)

for (i in names(df_list)) {
  df <- df_list[[i]]$df
  pollutant <- df_list[[i]]$pollutant
  temp <- df_list[[i]]$temp
  name <- df_list[[i]]$name
  
  df$date <- as.Date("2004-03-10") + df$dayssince
  df$week_start <- as.Date(cut(df$date, "week"))
  
  
  hist(df[[pollutant]], 
       xlab = paste(name, "Concentration (µg/m³)"), 
       col = "steelblue",
       main = paste("Distribution of", name),
       breaks = 20)
  
  boxplot(df[[pollutant]] ~ df$week_start,
          xlab = "",
          ylab = paste(name, "Concentration (µg/m³)"),
          main = paste("Weekly", name, "Distribution"),
          col = "steelblue",
          las = 2,
          cex.axis = 0.7)
  title(xlab = "Week Starting Date", line = 5)
  
  hist(df[[temp]], 
       xlab = "Temperature (°C)", 
       col = "coral",
       main = paste("Distribution of Temperature (", name, ")", sep = ""),
       breaks = 20)
  
  boxplot(df[[temp]] ~ df$week_start,
          xlab = "",
          ylab = "Temperature (°C)",
          main = paste("Weekly Temperature Distribution (", name, ")", sep = ""),
          col = "coral",
          las = 2,
          cex.axis = 0.7)
  title(xlab = "Week Starting Date", line = 5)
  
}
 
```


The following plots show frequencies of observations as well as weekly differences in temperature and particle concentrations for CO and NOx. 

The distribution of CO is indeed very left-skewed with most observations being around 1 or 2 µg/m3. From around a concentration of 1, the frequency goes downwards exponential-like. Very few observations are in the  upper half of the range. 

The weekly CO distribution have only few and quite small trends. The IQR is relatively even in most weeks, however there seems to be a slight spike around the winter months. This is probably due to more people heating up their houses or choosing to move by car in these months. The plot also shows a relatively high number of outliers. 

The temperature distribution (CO) is relatively symmetrical and doesn't look too far from a normal distribution. As described in 1), it is however slightly left-skewed. The most frequent tempatures are between 12 and 14 degrees. 

The weekly temperature distribution (CO) shows very few anomalies. Most of the weekly box plots have no outliers, and each box plot has a relatively small ranges, especially considering observations were made during both night- and daytime. The box plots show a general pattern of what you would expect from temperatures across a year with four seasons in Italy. 

The frequency distribution of NOx looks very similar to that of CO, however, as mentioned in 1), with a significantly bigger range. A very left-skewed distribution indeed. 

This looks roughly similar to the weekly CO distribution, but with a more significant spike during the winter months. Also, the spike comes quite suddenly around the beginning of september, and the medians of the box plots don't decrease significantly again for the spring of 2025. We can't seem to find a good cause for this. The volcano of Etna erupted quite exactly when the sudden increase in NOx-level comes, but it shouldn't have made a noteworthy difference.
Similarly to CO, there are a lot of outliers in the weekly distribution.

The frequency distribution and grouped box plot for temperature (NOx) is almost identical to those of CO for the reason described in 1) and thus won't be discussed further.

#  3. Outlier detection

We are going to identify if there might be any potential outliers within the CO concentration dataset. This could be CO concentrations that are drastically high or low to the point where the measurements differ significantly from the rest of the dataset. The threshold for what measurements can be considered outliers or not is defined by the following:
Q1 - 1.5 * IQR
Q3 + 1.5 * IQR
The lower bound is defined as the first quartile subtracted by 1.5 times the IQR, inter quantile range, which is the difference between the first and third quartiles. The upper bound is calculated similarly, defined as the third quartile plus 1.5 times the IQR.

The boxplot function in R uses this method to identify outliers when generating a box plot. We will therefore use this function to identify any potential outliers within the CO concentration dataset. The box plot that also gets generated will be helpful for analyzing the outliers after they’ve been identified.
  
```{r}

boxplot(dataframeco$co,horizontal = TRUE,
        main = "CO Concentration Distribution",
        xlab = "CO Concentration (µg/m³)",
        ylab = "",
        col = "steelblue",
        border = "darkblue",
        notch = TRUE,
        outpch = 16,
        outcol = "red",
        outcex = 1)

boxplot.stats(dataframeco$co)

```

R has identified a number of outliers in the dataset, all exceeding the upper threshold explained earlier. There is a total of 257 outliers, with values ranging from 5.4 to 11.9 µg/m3. These values are quite high considering the median value is 1.8 and the third quartile is 2.8. Looking at both the boxplot above, and the “Distribution of CO” bar chart from section 2, the outliers appear to be distributed quite closely following a curve, with higher values being increasingly less common. This non-irregular distribution resembles the distribution of the rest of the dataset quite closely.

These outlier measurements could be caused by very local releases of CO, such as a plume of smoke from a vehicle, fire or other source. The relatively non-irregular pattern of outliers might, however, contradict this being the source of outlier measurements. Since the measurements were taken in an Italian city, it is also possible this city is located in a valley, where certain weather conditions could trap the pollution of the city and prevent the CO particles from dispersing leading to high measurements. We do, however, not know for sure what has caused these outliers. 

Since the outlier measurements are distributed in a non-irregular manner which conforms relatively well to the distribution pattern of the rest of the dataset, these may not be “true” outliers in the sense that they do not differ significantly from the rest of the dataset. These outliers have simply been flagged as such via the IQR method on account of the values being relatively high. They are therefore worth keeping for further analysis.

#  4. Correlation analysis

```{r}


cor(dataframeco$time,dataframeco$co)

plot(dataframeco$time, dataframeco$cogt,
     main = "CO Concentration vs Time of Day",
     xlab = "Time of Day (hours)",
     ylab = "CO Concentration (µg/m³)",
     col = "steelblue",
     pch = 16,  # Solid circles
     cex = 0.5,  # Point size
     cex.main = 1.2,  # Title size
     cex.lab = 1.1)  # Label size

# Add grid
grid(col = "lightgray", lty = "dotted")


```

The scatterplot shows the relationship between the time of day (in hours) and the CO concentration. The CO concentration is measured in micrograms per cubic meter (µg/m³)
The correlation coefficient is approximately 0.36, which indicates a moderate positive relationship between CO concentration and time of day. While there is a general upward trend, the substantial dispersion of the data points suggests the relationship is not strong.

The scatter plot exhibits two peaks throughout the day, the first smaller peak in the morning rush hour, specifically around 8-9, the concentration then decreases and remains relatively stable from approximately 10 to 16. The major peak then arrives around 18-19, following the second rush hour of the day. From 20 till 6 the CO concentration has a downwards trend and drops every hour. The time of day with the lowest concentration of CO is around 4 till 6 during the night.

This pattern is consistent with typical daily traffic activity. The peaks in CO concentration during the morning (around 8–9) and evening (around 18–19) correspond to rush hours, when traffic volume and vehicle emissions are highest.

Between these periods, the concentration remains relatively stable, reflecting moderate but steady daytime traffic. The lowest concentrations occur during the night, especially between 4 and 6, when most people are asleep and traffic is minimal, leading to reduced emissions. 
While traffic patterns explain much of the observed variation, other factors such as weather conditions (wind speed and temperature) may also influence CO concentrations.





#  5. Statistical testing

```{r}
rhdataframeco <- data.frame(
  id = dataframeco$id,
  date = dataframeco$date,
  dayssince = dataframeco$dayssince,
  time = dataframeco$time,
  co = dataframeco$co,
  temperature = dataframeco$temperature
)

for (x in 0:24) {
  if ((x >= 7 && x <= 9) || (x >= 16 && x <= 18)) {
  } else {
 rhdataframeco <- rhdataframeco[rhdataframeco$time != x, ] 
  }
} 

ohdataframeco <- data.frame(
  id = dataframeco$id,
  date = dataframeco$date,
  dayssince = dataframeco$dayssince,
  time = dataframeco$time,
  co = dataframeco$co,
  temperature = dataframeco$temperature
)

for (x in 0:24) {
  if (x >= 22  || x <= 5) {
  } else {
 ohdataframeco <- ohdataframeco[ohdataframeco$time != x, ] 
  }
} 

t.test(rhdataframeco$co, ohdataframeco$co, paired = FALSE)
```
The computed p-value of 2,2*10-16 shows that there is a significant difference between rush hour and off hour. The median value of Rh is 2,67 ppm and 1,34 at Oh, this gives at different of 1,33 ppm, which is significant, almost double the Oh concentration.
This difference suggests that the concentration of CO in the highly linked to traffic amounts, which makes sense do to the fact that CO is a product of the burning of gasoline in combustion engines. 
It can their bye be concluded that the amount of traffic is a significant contributing factor of the concentration of CO. 

#  6. Regression analysis

```{r}

model_0 <- lm(no2gt ~ time, data = dataframeno2)
summary(model_0)

model_1 <- lm(no2gt ~ time + temperature, data = dataframeno2)
summary(model_1)

anova(model_0,model_1)

scenario <- data.frame(time = 9, temperature = 14)
predict(model_1, newdata = scenario, interval = "prediction")

residuals <- residuals(model_1)
qqnorm(residuals)
qqline(residuals)
       
```

Using the command ”lm”, you can create a simple linear regression, with the no2 as our data, and time as our explanatory variable. The resulting linear regression (model_0) becomes no2 =81.67+2.58·time, where 81.67 is the predicted levels of no2 when the time is 0, and 2.58 being the rate that the no2 increases every hour.
	
Same thing as the linear regression before, but where time is an added variable, which is added as another explanatory variable. The resulting linear regression (model:1) is now no2 =102.92+2.94·time-1.42·temperature, where our no2 now is predicted to be 102.92, where both the time and temperature is 0.

The model (model_1) now has 3 coefficients. Intercept, time and temperature. Our intercept is the value where both the temperature and time are equal to 0. This is the baseline for the linear regression. The time coefficient is the number of units the no2 increases or decreases per hour, no2 in our case it increases with 2.94 no2 The temperature coefficient is how much no2 increases or decreases by each extra degree, in this case it decreases by -1.42 no2.

The model (model_1) can be assessed using the values given to us by the “anora” command and comparing the two models. The first thing that indicates an improvement is the r2 value has increased from 0.137 to 0.207, which shows that the addition of temperature explains more of the variation. The second indication is that the second model has a decrease of “residual standard error”, from 44.25 to 42.43. The third and biggest indication, is the P value given to the second model by the “anora” command. It shows that the p value is 2.773102e-137 which means that the result of the model is very unlikely to be caused by chance, which makes the variable statistically significant.

Using the “predict” command, the model will predict the levels of no2, with a specific value set to the explanatory variables. In this example the time is set to 9 and our temperature is set to 14. Using those variables, the model predicted the no2 to be 109.46, with the predicted intervals to be [26.28,192.65]




#  7. Time series analysis and forecasting

```{r}

dataframenox$date <- as.Date("2004-03-10") + dataframenox$dayssince

daily_nox <- aggregate(noxgt ~ date, data = dataframenox, FUN = mean)

plot(daily_nox$date, daily_nox$noxgt, 
     type = "l", 
     col = "steelblue",
     xlab = "Days Since Start",
     ylab = "NOx Concentration (µg/m³)",
     main = "Daily Average NOx Concentration Over Time",
     lwd = 1.5)
lines(lowess(daily_nox$date, daily_nox$noxgt), col = "red", lwd = 2)
grid()

legend("topright", 
      legend = c("Daily NOx", "Trend"), 
      col = c("steelblue", "red"), 
      lwd = c(1.5, 2))

nox_ts <- ts(daily_nox$noxgt, frequency = 7)

EXP_model <- HoltWinters(nox_ts)

predicted_7day <- predict(EXP_model, n.ahead = 7)

last_actual_date <- max(daily_nox$date)
forecast_dates <- seq(last_actual_date + 1, by = "day", length.out = 7)

plot(daily_nox$date, daily_nox$noxgt, 
     type = "l", 
     col = "steelblue",
     xlab = "Date", 
     ylab = "NOx Concentration (µg/m³)",
     main = "NOx Time Series with 7-Day Forecast",
     xlim = c(min(daily_nox$date), max(daily_nox$date) + 14),
     ylim = range(c(daily_nox$noxgt, predicted_7day), na.rm = TRUE))

lines(forecast_dates, predicted_7day, col = "red", lwd = 2)

legend("topright", 
       legend = c("Observed", "Forecast"), 
       col = c("steelblue", "red"), 
       lty = 1, 
       lwd = 2)

grid()

               
```

The model, “Daily Average NOx Concentration Over Time”, shows the NOx pollution over several months and the trend the model is following, through the LOWESS line. The model shows some variability with clear fluctuations in NOx levels. The LOWESS trend suggests a slight upwards trend over time. There is not really a seasonal pattern.

The model is then changed to make a 7-day forecast to the Nox levels, via the “Holt-Winters” command, and using a 7-day frequency. The result of this model, "NOx Time Series with 7-Day Forecast", we can see the predicted forecast after 7 days, which is seen in the red part. It shows that the Nox levels will keep increasing through the week, until the middle of the week where, it falls back to a bit more than the starting value. 

The 7 day forecast is decently reliable for a short-term prediction, which is helped by the exponential smoothing, aka the “Holt-Winters” command . However, the model assumes that the trends and patterns will continue. It doesn’t account for external factors such as weather and traffic changes. That can significantly influence NOx levels, which won’t be accounted for.




