---
title: "Air Quality Exam Project"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# 0. Data Preparation

Loading of the Data
```{r}
data <- read.table("AirQualityUCI.csv", header = TRUE, sep = ";",dec = ",")

```

Construct a dataframe with relevant data.
```{r}
dataframeco <- data.frame(
  id = c(0:9470),
  date = data$Date,
  dayssince = c(0:0),
  time = data$Time,
  cogt = data$CO.GT,
  temperature = data$T
)

dataframenox <- data.frame(
  id = c(0:9470),
  date = data$Date,
  dayssince = c(0:0),
  time = data$Time,
  noxgt = data$NOx.GT.,
  temperature = data$T
)
```

Remove all the data with Not Assigned values.

```{r}
 dataframeco <- dataframeco[dataframeco$time >= 0, ] 
 dataframenox <- dataframenox[dataframenox$time >= 0, ] 
 dataframeco <- dataframeco[dataframeco$cogt != -200.0, ]
 dataframenox <- dataframenox[dataframenox$noxgt != -200.0, ]
 dataframeco <- dataframeco[dataframeco$temperature != -200.0, ]
 dataframenox <- dataframenox[dataframenox$temperature != -200.0, ]
```

Derive Additional Variable "Days Since". That will denote days since data observation begun

```{r}
 date1string <- "10/03/2004"
   date1 <-  strptime(as.character(date1string), "%d/%m/%Y")
   results <- numeric(nrow(dataframeco))
   for (i in 1:nrow(dataframeco)) {
     date2string <- dataframeco$date[i]
     date2 <-  strptime(as.character(date2string), "%d/%m/%Y")
     results[i] <-  difftime(date2,date1, units = "weeks")
   }
   dataframeco$dayssince <- round(results, digits = 0)
   
   results <- numeric(nrow(dataframenox))
   for (i in 1:nrow(dataframenox)) {
     date2string <- dataframenox$date[i]
     date2 <-  strptime(as.character(date2string), "%d/%m/%Y")
     results[i] <-  difftime(date2,date1, units = "weeks")
   }
   dataframenox$dayssince <- round(results, digits = 0)
```

Restructure time variable

```{r}
dataframeco$time <- gsub(".00.00", "", dataframeco$time)
dataframeco$time <- as.numeric(dataframeco$time)
dataframenox$time <- gsub(".00.00", "", dataframenox$time)
dataframenox$time <- as.numeric(dataframenox$time)

```
  
Update Id  

```{r}
    ids <- numeric(nrow(dataframeco))
    x <- -1
    for (i in 1:nrow(dataframeco)) {
      x = x + 1
      ids[i] <-  x
    }
    dataframeco$id <- ids
    
    ids <- numeric(nrow(dataframenox))
    y <- -1
    for (i in 1:nrow(dataframenox)) {
      y = y + 1
      ids[i] <-  y
    }
    dataframenox$id <- ids
```

 
#  1. Quantitative Exploratory Analysis
  
  Selected Variables: NOx (Kvælstofoxider), CO (Carbon Monoxide) Temperature
   
```{r}

summary(dataframenox$nox)

summary(dataframeco$co)

summary(dataframenox$temperature)

summary(dataframeco$temperature)



```

#  2. Visual Exploratory Analysis


```{r}

hist(dataframeco$co)

boxplot(dataframeco$co ~ dataframeco$dayssince, xlab = "Weeks since Start")

hist(dataframenox$nox)

boxplot(dataframenox$nox ~ dataframenox$dayssince, xlab = "Weeks since Start")

hist(dataframenox$dayssince)

boxplot(dataframenox$temperature ~ dataframenox$dayssince, xlab = "Weeks since Start")

hist(dataframeco$temperature)

boxplot(dataframeco$temperature ~ dataframeco$dayssince, xlab = "Weeks since Start")
 


```

#  3. Outlier detection
  
  
```{r}
  boxplot(dataframeco$co)
  boxplot.stats(dataframeco$co)
```

#  4. Correlation analysis

```{r}


cor(dataframeco$time,dataframeco$co)

plot(dataframeco$time,dataframeco$co)

```

#  5. Statistical testing

```{r}


rhdataframeco <- data.frame(
  id = dataframeco$id,
  date = dataframeco$date,
  dayssince = dataframeco$dayssince,
  time = dataframeco$time,
  co = dataframeco$co,
  temperature = dataframeco$temperature
)

rhdataframeco <- rhdataframeco[rhdataframeco$time >= 7, ]
rhdataframeco <- rhdataframeco[rhdataframeco$time <= 18, ]
rhdataframeco <- rhdataframeco[rhdataframeco$time != 10, ]
rhdataframeco <- rhdataframeco[rhdataframeco$time != 11, ]
rhdataframeco <- rhdataframeco[rhdataframeco$time != 12, ]
rhdataframeco <- rhdataframeco[rhdataframeco$time != 13, ]
rhdataframeco <- rhdataframeco[rhdataframeco$time != 14, ]
rhdataframeco <- rhdataframeco[rhdataframeco$time != 15, ]

ohdataframeco <- data.frame(
  id = dataframeco$id,
  date = dataframeco$date,
  dayssince = dataframeco$dayssince,
  time = dataframeco$time,
  co = dataframeco$co,
  temperature = dataframeco$temperature
)

ohdataframeco <- ohdataframeco[ohdataframeco$time != 6, ]
ohdataframeco <- ohdataframeco[ohdataframeco$time != 7, ]
ohdataframeco <- ohdataframeco[ohdataframeco$time != 8, ]
ohdataframeco <- ohdataframeco[ohdataframeco$time != 9, ]
ohdataframeco <- ohdataframeco[ohdataframeco$time != 10, ]
ohdataframeco <- ohdataframeco[ohdataframeco$time != 11, ]
ohdataframeco <- ohdataframeco[ohdataframeco$time != 12, ]
ohdataframeco <- ohdataframeco[ohdataframeco$time != 13, ]
ohdataframeco <- ohdataframeco[ohdataframeco$time != 14, ]
ohdataframeco <- ohdataframeco[ohdataframeco$time != 15, ]
ohdataframeco <- ohdataframeco[ohdataframeco$time != 16, ]
ohdataframeco <- ohdataframeco[ohdataframeco$time != 17, ]
ohdataframeco <- ohdataframeco[ohdataframeco$time != 18, ]
ohdataframeco <- ohdataframeco[ohdataframeco$time != 19, ]
ohdataframeco <- ohdataframeco[ohdataframeco$time != 20, ]
ohdataframeco <- ohdataframeco[ohdataframeco$time != 21, ]

```