---
title: "Air Quality Exam Project"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# 0. Data Preparation

Loading of the Data
```{r}
data <- read.table("AirQualityUCI.csv", header = TRUE, sep = ";",dec = ",")

```

Construct dataframes with relevant data.
```{r}
dataframeco <- data.frame(
  id = c(0:9470),
  date = data$Date,
  dayssince = c(0:0),
  time = data$Time,
  cogt = data$CO.GT,
  temperature = data$T
)

dataframenox <- data.frame(
  id = c(0:9470),
  date = data$Date,
  dayssince = c(0:0),
  time = data$Time,
  noxgt = data$NOx.GT.,
  temperature = data$T
)

dataframeno2 <- data.frame(
  id = c(0:9470),
  date = data$Date,
  time = data$Time,
  no2gt = data$NO2.GT.,
  temperature = data$T
)
```

Remove all the data with Not Assigned values.

```{r}
 dataframeco <- dataframeco[dataframeco$time >= 0, ] 
 dataframenox <- dataframenox[dataframenox$time >= 0, ] 
 dataframeno2 <- dataframeno2[dataframeno2$time >= 0, ] 
 
 dataframeco <- dataframeco[dataframeco$cogt != -200.0, ]
 dataframenox <- dataframenox[dataframenox$noxgt != -200.0, ]
 dataframeno2 <- dataframeno2[dataframeno2$no2gt != -200.0, ]
 
 dataframeco <- dataframeco[dataframeco$temperature != -200.0, ]
 dataframenox <- dataframenox[dataframenox$temperature != -200.0, ]
 dataframeno2 <- dataframeno2[dataframeno2$temperature != -200.0, ]
```

Derive Additional Variable "Days Since". That will denote days since data observation begun

```{r}
 date1string <- "10/03/2004"
   date1 <-  strptime(as.character(date1string), "%d/%m/%Y")
   results <- numeric(nrow(dataframeco))
   for (i in 1:nrow(dataframeco)) {
     date2string <- dataframeco$date[i]
     date2 <-  strptime(as.character(date2string), "%d/%m/%Y")
     results[i] <-  difftime(date2,date1, units = "days")
   }
   dataframeco$dayssince <- round(results, digits = 0)
   
   results <- numeric(nrow(dataframenox))
   for (i in 1:nrow(dataframenox)) {
     date2string <- dataframenox$date[i]
     date2 <-  strptime(as.character(date2string), "%d/%m/%Y")
     results[i] <-  difftime(date2,date1, units = "days")
   }
   dataframenox$dayssince <- round(results, digits = 0)
```

Restructure time variable

```{r}
dataframeco$time <- gsub(".00.00", "", dataframeco$time)
dataframeco$time <- as.numeric(dataframeco$time)
dataframenox$time <- gsub(".00.00", "", dataframenox$time)
dataframenox$time <- as.numeric(dataframenox$time)
dataframeno2$time <- gsub(".00.00", "", dataframeno2$time)
dataframeno2$time <- as.numeric(dataframeno2$time)

```
  
Update Id  

```{r}
    ids <- numeric(nrow(dataframeco))
    x <- -1
    for (i in 1:nrow(dataframeco)) {
      x = x + 1
      ids[i] <-  x
    }
    dataframeco$id <- ids
    
    ids <- numeric(nrow(dataframenox))
    y <- -1
    for (i in 1:nrow(dataframenox)) {
      y = y + 1
      ids[i] <-  y
    }
    dataframenox$id <- ids
```

 
#  1. Quantitative Exploratory Analysis
  
  Selected Variables: NOx (Kvælstofoxider), CO (Carbon Monoxide) Temperature
   
```{r}

summary(dataframenox$nox)
var(dataframenox$nox)
sd(dataframenox$nox)

```

There range is very big. As the 3rd quantile is 321 and the max is 1479, there are relatively few observations in the more extremely high values. And as the median is lower than the mean that also means that less observations are in the lower half than in the upper one. The difference between minimum-1st quantile, 1st quantile-median, and median-3rd quantile are, however, all relatively similar. All in all, the distribution seems to be quite left-skewed (positive)

```{r}

summary(dataframeco$co)
var(dataframeco$co)
sd(dataframeco$co)

```

Interestingly a lot smaller range than for NOx. The situation with max is however very similar to that of Nox, as the difference between 3rd quantile and max is a lot bigger than all other quantile differences. The mean is also higher than the median. Again, this is a left-skewed (positive) distribution. 


```{r}
summary(dataframenox$temperature)
var(dataframenox$temperature)
sd(dataframenox$temperature)
```
This should be left-skewed again, but less obviously. The difference between 3rd quantile and maximum is still the biggest of all, but the range between minimum and 1st quantile is also quite big. Meaning, the distribution is still left-skewed, but a lot more symmetrical than for NOx and CO. Most observations are close to the median.

```{r}

summary(dataframeco$temperature)
var(dataframeco$temperature)
sd(dataframeco$temperature)

```


#  2. Visual Exploratory Analysis



```{r}

df_list <- list(
  co = list(df = dataframeco, pollutant = "cogt", temp = "temperature", name = "CO"),
  nox = list(df = dataframenox, pollutant = "noxgt", temp = "temperature", name = "NOx")
)

for (i in names(df_list)) {
  df <- df_list[[i]]$df
  pollutant <- df_list[[i]]$pollutant
  temp <- df_list[[i]]$temp
  name <- df_list[[i]]$name
  
  df$date <- as.Date("2004-03-10") + df$dayssince
  df$week_start <- as.Date(cut(df$date, "week"))
  
  
  hist(df[[pollutant]], 
       xlab = paste(name, "Concentration (µg/m³)"), 
       col = "steelblue",
       main = paste("Distribution of", name),
       breaks = 20)
  
  boxplot(df[[pollutant]] ~ df$week_start,
          xlab = "",
          ylab = paste(name, "Concentration (µg/m³)"),
          main = paste("Weekly", name, "Distribution"),
          col = "steelblue",
          las = 2,
          cex.axis = 0.7)
  title(xlab = "Week Starting Date", line = 5)
  
  hist(df[[temp]], 
       xlab = "Temperature (°C)", 
       col = "coral",
       main = paste("Distribution of Temperature (", name, ")", sep = ""),
       breaks = 20)
  
  boxplot(df[[temp]] ~ df$week_start,
          xlab = "",
          ylab = "Temperature (°C)",
          main = paste("Weekly Temperature Distribution (", name, ")", sep = ""),
          col = "coral",
          las = 2,
          cex.axis = 0.7)
  title(xlab = "Week Starting Date", line = 5)
  
}
 
```

#  3. Outlier detection

We are going to identify if there might be any potential outliers within the CO concentration dataset. This could be CO concentrations that are drastically high or low to the point where the measurements differ significantly from the rest of the dataset. The threshold for what measurements can be considered outliers or not is defined by the following:
Q1 - 1.5 * IQR
Q3 + 1.5 * IQR
The lower bound is defined as the first quartile subtracted by 1.5 times the IQR, inter quantile range, which is the difference between the first and third quartiles. The upper bound is calculated similarly, defined as the third quartile plus 1.5 times the IQR.

The boxplot function in R uses this method to identify outliers when generating a box plot. We will therefore use this function to identify any potential outliers within the CO concentration dataset. The box plot that also gets generated will be helpful for analyzing the outliers after they’ve been identified.
  
```{r}

boxplot(dataframeco$co,horizontal = TRUE,
        main = "CO Concentration Distribution",
        xlab = "CO Concentration (µg/m³)",
        ylab = "",
        col = "steelblue",
        border = "darkblue",
        notch = TRUE,
        outpch = 16,
        outcol = "red",
        outcex = 1)

boxplot.stats(dataframeco$co)

```

R has identified a number of outliers in the dataset, all exceeding the upper threshold explained earlier. There is a total of 257 outliers, with values ranging from 5.4 to 11.9 µg/m3. These values are quite high considering the median value is 1.8 and the third quartile is 2.8. Looking at both the boxplot above, and the “Distribution of CO” bar chart from section 2, the outliers appear to be distributed quite closely following a curve, with higher values being increasingly less common. This non-irregular distribution resembles the distribution of the rest of the dataset quite closely.

These outlier measurements could be caused by very local releases of CO, such as a plume of smoke from a vehicle, fire or other source. The relatively non-irregular pattern of outliers might, however, contradict this being the source of outlier measurements. Since the measurements were taken in an Italian city, it is also possible this city is located in a valley, where certain weather conditions could trap the pollution of the city and prevent the CO particles from dispersing leading to high measurements. We do, however, not know for sure what has caused these outliers. 

Since the outlier measurements are distributed in a non-irregular manner which conforms relatively well to the distribution pattern of the rest of the dataset, these may not be “true” outliers in the sense that they do not differ significantly from the rest of the dataset. These outliers have simply been flagged as such via the IQR method on account of the values being relatively high. They are therefore worth keeping for further analysis.

#  4. Correlation analysis

```{r}


cor(dataframeco$time,dataframeco$co)

plot(dataframeco$time, dataframeco$cogt,
     main = "CO Concentration vs Time of Day",
     xlab = "Time of Day (hours)",
     ylab = "CO Concentration (µg/m³)",
     col = "steelblue",
     pch = 16,  # Solid circles
     cex = 0.5,  # Point size
     cex.main = 1.2,  # Title size
     cex.lab = 1.1)  # Label size

# Add grid
grid(col = "lightgray", lty = "dotted")


```

#  5. Statistical testing

```{r}
rhdataframeco <- data.frame(
  id = dataframeco$id,
  date = dataframeco$date,
  dayssince = dataframeco$dayssince,
  time = dataframeco$time,
  co = dataframeco$co,
  temperature = dataframeco$temperature
)

for (x in 0:24) {
  if ((x >= 7 && x <= 9) || (x >= 16 && x <= 18)) {
  } else {
 rhdataframeco <- rhdataframeco[rhdataframeco$time != x, ] 
  }
} 

ohdataframeco <- data.frame(
  id = dataframeco$id,
  date = dataframeco$date,
  dayssince = dataframeco$dayssince,
  time = dataframeco$time,
  co = dataframeco$co,
  temperature = dataframeco$temperature
)

for (x in 0:24) {
  if (x >= 22  || x <= 5) {
  } else {
 ohdataframeco <- ohdataframeco[ohdataframeco$time != x, ] 
  }
} 

t.test(rhdataframeco$co, ohdataframeco$co, paired = FALSE)
```

#  6. Regression analysis

```{r}

model_0 <- lm(no2gt ~ time, data = dataframeno2)
summary(model_0)

model_1 <- lm(no2gt ~ time + temperature, data = dataframeno2)
summary(model_1)

anova(model_0,model_1)

scenario <- data.frame(time = 9, temperature = 14)
predict(model_1, newdata = scenario, interval = "prediction")

residuals <- residuals(model_1)
qqnorm(residuals)
qqline(residuals)
       



```

#  7. Time series analysis and forecasting

```{r}

dataframenox$date <- as.Date("2004-03-10") + dataframenox$dayssince

daily_nox <- aggregate(noxgt ~ date, data = dataframenox, FUN = mean)

plot(daily_nox$date, daily_nox$noxgt, 
     type = "l", 
     col = "steelblue",
     xlab = "Days Since Start",
     ylab = "NOx Concentration (µg/m³)",
     main = "Daily Average NOx Concentration Over Time",
     lwd = 1.5)
lines(lowess(daily_nox$date, daily_nox$noxgt), col = "red", lwd = 2)
grid()

legend("topright", 
      legend = c("Daily NOx", "Trend"), 
      col = c("steelblue", "red"), 
      lwd = c(1.5, 2))

nox_ts <- ts(daily_nox$noxgt, frequency = 7)

EXP_model <- HoltWinters(nox_ts)

predicted_7day <- predict(EXP_model, n.ahead = 7)

last_actual_date <- max(daily_nox$date)
forecast_dates <- seq(last_actual_date + 1, by = "day", length.out = 7)

plot(daily_nox$date, daily_nox$noxgt, 
     type = "l", 
     col = "steelblue",
     xlab = "Date", 
     ylab = "NOx Concentration (µg/m³)",
     main = "NOx Time Series with 7-Day Forecast",
     xlim = c(min(daily_nox$date), max(daily_nox$date) + 14),
     ylim = range(c(daily_nox$noxgt, predicted_7day), na.rm = TRUE))

lines(forecast_dates, predicted_7day, col = "red", lwd = 2)

legend("topright", 
       legend = c("Observed", "Forecast"), 
       col = c("steelblue", "red"), 
       lty = 1, 
       lwd = 2)

grid()

               
```





