---
title: "Air Quality Exam Project"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# 0. Data Preparation

Loading of the Data
```{r}
data <- read.table("AirQualityUCI.csv", header = TRUE, sep = ";",dec = ",")

```

Construct a dataframe with relevant data.
```{r}
dataframeco <- data.frame(
  id = c(0:9470),
  date = data$Date,
  dayssince = c(0:0),
  time = data$Time,
  cogt = data$CO.GT,
  temperature = data$T
)

dataframeno2 <- data.frame(
  id = c(0:9470),
  date = data$Date,
  dayssince = c(0:0),
  time = data$Time,
  no2gt = data$NO2.GT.,
  temperature = data$T
)
```

Remove all the data with Not Assigned values.

```{r}
 dataframeco <- dataframeco[dataframeco$time >= 0, ] 
 dataframeno2 <- dataframeno2[dataframeno2$time >= 0, ] 
 dataframeco <- dataframeco[dataframeco$cogt != -200.0, ]
 dataframeno2 <- dataframeno2[dataframeno2$no2gt != -200.0, ]
 dataframeco <- dataframeco[dataframeco$temperature != -200.0, ]
 dataframeno2 <- dataframeno2[dataframeno2$temperature != -200.0, ]
```

Derive Additional Variable "Days Since". That will denote days since data observation begun

```{r}
 date1string <- "10/03/2004"
   date1 <-  strptime(as.character(date1string), "%d/%m/%Y")
   results <- numeric(nrow(dataframeco))
   for (i in 1:nrow(dataframeco)) {
     date2string <- dataframeco$date[i]
     date2 <-  strptime(as.character(date2string), "%d/%m/%Y")
     results[i] <-  difftime(date2,date1, units = "days")
   }
   dataframeco$dayssince <- round(results, digits = 0)
   
   results <- numeric(nrow(dataframeno2))
   for (i in 1:nrow(dataframeno2)) {
     date2string <- dataframeno2$date[i]
     date2 <-  strptime(as.character(date2string), "%d/%m/%Y")
     results[i] <-  difftime(date2,date1, units = "days")
   }
   dataframeno2$dayssince <- round(results, digits = 0)
```

Restructure time variable

```{r}
dataframeco$time <- gsub(".00.00", "", dataframeco$time)
dataframeco$time <- as.numeric(dataframeco$time)
dataframeno2$time <- gsub(".00.00", "", dataframeno2$time)
dataframeno2$time <- as.numeric(dataframeno2$time)
```
  
Update Id  

```{r}
    ids <- numeric(nrow(dataframeco))
    x <- -1
    for (i in 1:nrow(dataframeco)) {
      x = x + 1
      ids[i] <-  x
    }
    dataframeco$id <- ids
    
    ids <- numeric(nrow(dataframeno2))
    y <- -1
    for (i in 1:nrow(dataframeno2)) {
      y = y + 1
      ids[i] <-  y
    }
    dataframeno2$id <- ids
```

 
#  1. Quantitative Exploratory Analysis
  
  Selected Variables: NO2 (Nitrogen Dioxide), CO (Carbon Monoxide), Days Since, Time and Temperature
   
```{r}

dataframeanal <- data.frame(
  name = c("no","co","dayssince","time","temperatureno","temperatureco"),
  min = c(min(dataframeno2$no2gt),min(dataframeco$cogt),0,0,min(dataframeno2$temperature),min(dataframeco$temperature)),
  #firstqu = c(quantile(dataframeno2$no2gt, probs = seq(0, 1, 0.25)),quantile(dataframeco$cogt, probs = seq(0, 1, 0.25)),2,3,quantile(dataframeno2$temperature, probs = seq(0, 1, 0.25)),quantile(dataframeco$temperature, probs = seq(0, 1, 0.25))),
  median = c(median(dataframeno2$no2gt),median(dataframeco$cogt),0,0,median(dataframeno2$temperature),median(dataframeco$temperature)),
  mean = c(mean(dataframeno2$no2gt),mean(dataframeco$cogt),0,0,mean(dataframeno2$temperature),mean(dataframeco$temperature)),
  thirdqu = c(0,1,2,3,4,5),
  max = c(max(dataframeno2$no2gt),max(dataframeco$cogt),0,0,max(dataframeno2$temperature),max(dataframeco$temperature))
)

  # print(min(dataframeno2$no2gt))
  #  variables <- c(dataframeno2$no2gt,dataframeco$cogt)
  # min <- numeric(nrow(dataframeqea))
  #   for (i in variables) {
  #     #print(i)
  #     #print(min[i]) 
  #     }
  #   dataframeqea$min <- min
  # 
  # #summary(dataframeco$time)

```

```{r}
  
  summary(dataframeno2$time)

```
  
  
  
```{r}

 agdata <- aggregate(dataframeco$cogt ~ dataframeco$dayssince, data = dataframeco, FUN = mean)
 #print(agdata)
 plot(agdata, col = "darkblue")
 
```

