---
title: "Air Quality Exam Project"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# 0. Data Preparation

Loading of the Data
```{r}
data <- read.table("AirQualityUCI.csv", header = TRUE, sep = ";",dec = ",")

```

Construct dataframes with relevant data.
```{r}
dataframeco <- data.frame(
  id = c(0:9470),
  date = data$Date,
  dayssince = c(0:0),
  time = data$Time,
  cogt = data$CO.GT,
  temperature = data$T
)

dataframenox <- data.frame(
  id = c(0:9470),
  date = data$Date,
  dayssince = c(0:0),
  time = data$Time,
  noxgt = data$NOx.GT.,
  temperature = data$T
)

dataframeno2 <- data.frame(
  id = c(0:9470),
  date = data$Date,
  time = data$Time,
  no2gt = data$NO2.GT.,
  temperature = data$T
)
```

Remove all the data with Not Assigned values.

```{r}
 dataframeco <- dataframeco[dataframeco$time >= 0, ] 
 dataframenox <- dataframenox[dataframenox$time >= 0, ] 
 dataframeno2 <- dataframeno2[dataframeno2$time >= 0, ] 
 
 dataframeco <- dataframeco[dataframeco$cogt != -200.0, ]
 dataframenox <- dataframenox[dataframenox$noxgt != -200.0, ]
 dataframeno2 <- dataframeno2[dataframeno2$no2gt != -200.0, ]
 
 dataframeco <- dataframeco[dataframeco$temperature != -200.0, ]
 dataframenox <- dataframenox[dataframenox$temperature != -200.0, ]
 dataframeno2 <- dataframeno2[dataframeno2$temperature != -200.0, ]
```

Derive Additional Variable "Days Since". That will denote days since data observation begun

```{r}
 date1string <- "10/03/2004"
   date1 <-  strptime(as.character(date1string), "%d/%m/%Y")
   results <- numeric(nrow(dataframeco))
   for (i in 1:nrow(dataframeco)) {
     date2string <- dataframeco$date[i]
     date2 <-  strptime(as.character(date2string), "%d/%m/%Y")
     results[i] <-  difftime(date2,date1, units = "days")
   }
   dataframeco$dayssince <- round(results, digits = 0)
   
   results <- numeric(nrow(dataframenox))
   for (i in 1:nrow(dataframenox)) {
     date2string <- dataframenox$date[i]
     date2 <-  strptime(as.character(date2string), "%d/%m/%Y")
     results[i] <-  difftime(date2,date1, units = "days")
   }
   dataframenox$dayssince <- round(results, digits = 0)
```

Restructure time variable

```{r}
dataframeco$time <- gsub(".00.00", "", dataframeco$time)
dataframeco$time <- as.numeric(dataframeco$time)
dataframenox$time <- gsub(".00.00", "", dataframenox$time)
dataframenox$time <- as.numeric(dataframenox$time)
dataframeno2$time <- gsub(".00.00", "", dataframeno2$time)
dataframeno2$time <- as.numeric(dataframeno2$time)

```
  
Update Id  

```{r}
    ids <- numeric(nrow(dataframeco))
    x <- -1
    for (i in 1:nrow(dataframeco)) {
      x = x + 1
      ids[i] <-  x
    }
    dataframeco$id <- ids
    
    ids <- numeric(nrow(dataframenox))
    y <- -1
    for (i in 1:nrow(dataframenox)) {
      y = y + 1
      ids[i] <-  y
    }
    dataframenox$id <- ids
```

 
#  1. Quantitative Exploratory Analysis
  
  Selected Variables: NOx (Kvælstofoxider), CO (Carbon Monoxide) Temperature
   
```{r}

summary(dataframenox$nox)

summary(dataframeco$co)

summary(dataframenox$temperature)

summary(dataframeco$temperature)



```

#  2. Visual Exploratory Analysis


```{r}

df_list <- list(
  co = list(df = dataframeco, pollutant = "cogt", temp = "temperature", name = "CO"),
  nox = list(df = dataframenox, pollutant = "noxgt", temp = "temperature", name = "NOx")
)

for (i in names(df_list)) {
  df <- df_list[[i]]$df
  pollutant <- df_list[[i]]$pollutant
  temp <- df_list[[i]]$temp
  name <- df_list[[i]]$name
  
  df$date <- as.Date("2004-03-10") + df$dayssince
  df$week_start <- as.Date(cut(df$date, "week"))
  
  
  hist(df[[pollutant]], 
       xlab = paste(name, "Concentration (µg/m³)"), 
       col = "steelblue",
       main = paste("Distribution of", name),
       breaks = 20)
  
  boxplot(df[[pollutant]] ~ df$week_start,
          xlab = "",
          ylab = paste(name, "Concentration (µg/m³)"),
          main = paste("Weekly", name, "Distribution"),
          col = "steelblue",
          las = 2,
          cex.axis = 0.7)
  title(xlab = "Week Starting Date", line = 5)
  
  hist(df[[temp]], 
       xlab = "Temperature (°C)", 
       col = "coral",
       main = paste("Distribution of Temperature (", name, ")", sep = ""),
       breaks = 20)
  
  boxplot(df[[temp]] ~ df$week_start,
          xlab = "",
          ylab = "Temperature (°C)",
          main = paste("Weekly Temperature Distribution (", name, ")", sep = ""),
          col = "coral",
          las = 2,
          cex.axis = 0.7)
  title(xlab = "Week Starting Date", line = 5)
  
}
 
```

#  3. Outlier detection
  
  
```{r}

boxplot(dataframeco$co,horizontal = TRUE,
        main = "CO Concentration Distribution",
        xlab = "CO Concentration (µg/m³)",
        ylab = "",
        col = "steelblue",
        border = "darkblue",
        notch = TRUE,
        outpch = 16,
        outcol = "red",
        outcex = 1)

boxplot.stats(dataframeco$co)

```

#  4. Correlation analysis

```{r}


cor(dataframeco$time,dataframeco$co)

plot(dataframeco$time,dataframeco$co)

```

#  5. Statistical testing

```{r}
rhdataframeco <- data.frame(
  id = dataframeco$id,
  date = dataframeco$date,
  dayssince = dataframeco$dayssince,
  time = dataframeco$time,
  co = dataframeco$co,
  temperature = dataframeco$temperature
)

for (x in 0:24) {
  if ((x >= 7 && x <= 9) || (x >= 16 && x <= 18)) {
  } else {
 rhdataframeco <- rhdataframeco[rhdataframeco$time != x, ] 
  }
} 

ohdataframeco <- data.frame(
  id = dataframeco$id,
  date = dataframeco$date,
  dayssince = dataframeco$dayssince,
  time = dataframeco$time,
  co = dataframeco$co,
  temperature = dataframeco$temperature
)

for (x in 0:24) {
  if (x >= 22  || x <= 5) {
  } else {
 ohdataframeco <- ohdataframeco[ohdataframeco$time != x, ] 
  }
} 

t.test(rhdataframeco$co, ohdataframeco$co, paired = FALSE)
```

#  6. Regression analysis

```{r}

model_0 <- lm(no2gt ~ time, data = dataframeno2)
summary(model_0)

model_1 <- lm(no2gt ~ time + temperature, data = dataframeno2)
summary(model_1)

anova(model_0,model_1)

scenario <- data.frame(time = 9, temperature = 14)
predict(model_1, newdata = scenario, interval = "prediction")

residuals <- residuals(model_1)
qqnorm(residuals)
qqline(residuals)
       
```

#  7. Time series analysis and forecasting

```{r}

dataframenox$date <- as.Date("2004-03-10") + dataframenox$dayssince

daily_nox <- aggregate(noxgt ~ date, data = dataframenox, FUN = mean)

plot(daily_nox$date, daily_nox$noxgt, 
     type = "l", 
     col = "steelblue",
     xlab = "Days Since Start",
     ylab = "NOx Concentration (µg/m³)",
     main = "Daily Average NOx Concentration Over Time",
     lwd = 1.5)
lines(lowess(daily_nox$date, daily_nox$noxgt), col = "red", lwd = 2)
grid()

legend("topright", 
      legend = c("Daily NOx", "Trend"), 
      col = c("steelblue", "red"), 
      lwd = c(1.5, 2))

nox_ts <- ts(daily_nox$noxgt, frequency = 7)

EXP_model <- HoltWinters(nox_ts)

predicted_7day <- predict(EXP_model, n.ahead = 7)

last_actual_date <- max(daily_nox$date)
forecast_dates <- seq(last_actual_date + 1, by = "day", length.out = 7)

plot(daily_nox$date, daily_nox$noxgt, 
     type = "l", 
     col = "steelblue",
     xlab = "Date", 
     ylab = "NOx Concentration (µg/m³)",
     main = "NOx Time Series with 7-Day Forecast",
     xlim = c(min(daily_nox$date), max(daily_nox$date) + 14),
     ylim = range(c(daily_nox$noxgt, predicted_7day), na.rm = TRUE))

lines(forecast_dates, predicted_7day, col = "red", lwd = 2)

legend("topright", 
       legend = c("Observed", "Forecast"), 
       col = c("steelblue", "red"), 
       lty = 1, 
       lwd = 2)

grid()

               
```





